---
title: "BIO331 -- Lab 07: Bioinformatics and Working with Strings of Biological Data"
subtitle: "Bioinformatics"
author: Dee Ruttenberg
date: Mar. 18, 2026
output: html_document
---

Today we're going to apply our work on R to Bioinformatic data sets. The typical output of a DNA sequencing experiment is a fastq file, consisting of often millions of configs which need to be stitched together (do you remember the two ways we do that?). Fastq files are different from fasta files as they contain quality scores -- this is useful to see how reliable your fastq data is.  The first tool we are going to use is Fastqc, to interpret the results of our fastq files.

# Quality Control

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("Rqc")
```

We will use a specific small dataset used specifically for teaching bioinformatics. Rqc will output an .html file (essentially a web page).  For this class, I will ask that you understand specifically the results of 'Average Quality', 'Cycle-Specific Average Quality', 'Cycle-specific Quality Distribution', and 'Cycle-Specific Base Call Proportion'.

```{r}
library(Rqc)
folder <- system.file(package="ShortRead", "extdata/E-MTAB-1147")
rqc(path = folder, pattern = ".fastq.gz")
```

# Bioconductor

Bioconductor and Biostrings are especially powerful tools to examine fasta files in R.  In this section, we'll learn some tools to manipulate them.  

```{r}
library(Biostrings)
seqs <- readDNAStringSet("Full_cmv_259sequences_Aln.fasta")
```

As Fasta files are labeled, you can identify specific sequences, either by loading a specfic label or identifying a keyword:

```{r}
my_seq <- seqs["Europe_United Kingdom_GQ221974.1_Human_herpesvirus_5_strain_3157_complete_genome"]
names_with_uk <- grep("United Kingdom", names(seqs), ignore.case = TRUE)
UK_seqs <- seqs[names_with_europe]
UK_subset <- subseq(UK_seqs, start = 20000, end = 25000)
```

You can, using R, code an aligned genome as a matrix, in order to identify snps. 

```{r}
mat <- as.matrix(UK_seqs)
snp_positions <- which(apply(mat, 2, function(col) {
  bases <- col[col %in% c("A","C","G","T")]
  length(unique(bases)) > 1
}))
snp_matrix <- mat[, snp_positions]
snp_list <- lapply(snp_positions, function(pos) {
  data.frame(
    Sequence = rownames(mat),
    Position = pos,
    Base = mat[, pos]
  )
})
names(snp_list) <- paste0("Pos_", snp_positions)
```

#Alignment

The final skill we need to learn is how to align .fasta files to a reference genome.  The data we have today is aligned so we'll have to arbitrarily "unalign" it.  We will use our first sequence as the reference.  

```{r}
UK_seqs_nogap <- DNAStringSet(gsub("-", "", as.character(UK_seqs))) 
ref <- seqs[6]
```

The region of interest from hcmv will be our query site.   

```{r}
library(Biostrings)
query <- readDNAStringSet("example.fasta")
ref <- seqs["Europe_Germany_JX512204.1_Human_herpesvirus_5_strain_HAN16_complete_genome"]
#ref <- readDNAStringSet("hcmvgenome.fasta")
query <- DNAString(toupper(as.character(query)))
ref <- DNAStringSet(gsub("-", "", as.character(ref))) 

hits <- matchPattern(query, ref[[1]], max.mismatch=20)

```

We can now look at our aligned data!

```{r}
alignments
```

# Homework Question 1: Interpretting QC data.

a. Look at the 'Cycle-specific Base Call Proportion' rqc file.  What is the ratio of A/C/G/T in cycles 1-10. What is the ratio of A/C/G/T in cycles 65-75?  Based on your knowledge of DNA-sequencing, propose a hypothesis for why the graphs  have a different ratio.
Answer: Based on the graphs titled " Cycle-specific GC content, and Cycle-specific quality distribution" I was only able to see the percent's of GC. Looking at the first graph titled cycle-specific GC content, cycles 1-10 are not consistent, or don't have a consistent trend that I can interpret clearly, this might be noisy data, and there is a lot of variation. However, for cycles 65-75 I could see that there is an overall trend, and that the GC content is around 52-56%. The average of 52 and 56 is 54, when I divide that by 2 I get 27, so the ratio for that would be 27G/27C. In order to get A/T ratio I just subtracted 100 from 54 and I got 46. When I divide that by 2 I got 23A/23T so the full ratio is 23/27/27/23. After looking at the graph titled" cycle-specific Base call proportion graph, this was confirmed. #Based on my knowledge of DNA- sequencing, a hypothesis for why the graphs have a different ratio is that models make assumptions leading to limitations. There are many experimental protocols that have some biases as well. For example, PCR amplifications can create biases in the sequence library and some sequences may be over-represented compared to others due to difficulties encountered during sequencing. 
b. Look at the 'Cycle specific Quality Distribution'. What is the typial quality of cycles 1-10?  What is the typical quality of cycles 65-75?  Based on your knowledge of DNA-sequencing, propose a hypothesis for why the cycles have a different quality
 Based on the graph, the typical quality of cycles 1-10 aare about 30-40 or 35-40 because it's mostly dark blue. The typical quality of cycles 65-75 is about 15-20, I say this because, it does have more orange color, but there is still some dark blue as well. Based on my knowledge of DNA-sequencing, a hypothesis for why the cycles have a different quality is because sequencing methods have some biases and errors like discussed in our lecture slides. Also, there could be more errors overtime which can lead to the cycles having a different quality.

# Homework Question 2: Loading a File, Reverse Complements

For one of the research projects, we will be studying a specific protein coding site of hcmv associated with the transition of hcmv from latent to active. Align the following two queries to any genome in the reference file. Which one successfully aligns? What does that mean about the site of interest?
2 answer: The one that succesfully aligns is, the original query sequence. I was able to figure this out becuase when I ran the code, i got a hit at the positions labeled start at 111721 and ended at 11200, there was definetly a match that was found. This means that the site of interest is in the reference genome, like the same layout of the orginal sequence. 
```{r}
query <- readDNAStringSet("hcmv15-hcmv1b-1.fasta")
reversecompquery <- reverseComplement(query)
```
# Homework Question 3: Isolating a Site

Isolate the 'hcmv15-hcmv1b-1.fasta' site in all 259 seqs. Your final result should be a DNAString set of 259 sequences about 4-5 kb long.  